---
title: "Changes in Geographic Dispersion of Mexican Migrants in the United States"
author: "Neal Marquez"
institute: "University of Washington"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

# Background

- Declining Population of Mexican Migrants in the US
- Shifting geographic distributions of Mexican migrants
- Aging population of Mexican migrants
- Potential labor force participation shifts

---

# Review on Theories of New Destinations

- Saturation of the job market in traditional destinations
- Change in political/social attitude toward migrants in traditional destinations
- Labor demands in new locations
- Changes in the border patrolling
- Changes in the demographics of migrants

---

# Characteristics of New Migrant Locations

- New source locations
- Younger population than the traditional location migrants
- New markets in multiple industries

---

# Age Comparisons of Migrants

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.height=4, dev='svg')
```

```{r include=FALSE}
library(ggplot2)
histogramNat <- readRDS("../../results/histogramNational.Rds")
ipumsPlot <- readRDS("../../results/ipumsPlot.Rds")

```


```{r}
histogramNat$mexRegional +
  theme(strip.text.y = element_text(size=4.5))
```

---

# Regional Comparisons to Native Population

```{r}
histogramNat$regionalCompare +
  theme(strip.text.y = element_text(size=4.5))
```

---

# Research Questions

- How have changes in the demographics of Mexican migrants altered geographic distributions
- What are the driving factors of movement in an aging population
    - Most theoretical models deal with young populations
- How does this aging/relocating effect labor force participation
    
---

# Shifting Age structure

```{r}
ipumsPlot$hist1 +
  theme(strip.text.y = element_text(size=5))
```

---

# Shifting Age structure

```{r}
ipumsPlot$hist2 +
  theme(strip.text.y = element_text(size=5))
```

---

# Shifting Age structure

```{r}
ipumsPlot$hist3 +
  theme(strip.text.y = element_text(size=5))
```

---

# Declining West Coast Population

```{r}
ipumsPlot$totalPopGrowth
```

---

# Declining West Coast Population

```{r}
ipumsPlot$hispPopGrowth
```

---
# Population Decline Correlations

```{r}
ipumsPlot$otherMexGrow
```

---

# Population Decline Correlations

```{r}
ipumsPlot$chicMexGrow
```

---

# Population Decline Correlations

```{r}
ipumsPlot$mexStatePeak +
  theme_void()
```

---

# Next Steps

- Check how changes in migrant populations alter age structure
- Adjust for expectations given mortality rates of Hispanics
- Try and find data on location pathways (stepwise migration)
- How does LFP by sector shift with increases in migrant population
- Test the relationship between housing market and labor sector shifts and migrant increases
